Numerical simulations of the Cordilleran ice sheet through the last glacial cycle

### 2015-06-19 cordillera-cycle.tex
Add corrections from Irina (part 2).

% Running title
\runningtitle{Climate forcing for Cordilleran ice sheet simulations}

@@

% =============================================================================
\section{Comparison to the geologic record}
\label{sec:discussion}
% =============================================================================

Large variations in the model response reveal its sensitivity to evolving
climate forcing. Nevertheless, geological evidence of former glaciations
provide a basis for validation of our runs, while the results from numerical
modelling can perhaps help to analyse some of the complexity of this evidence.
In this section, we compare model output to the geologic record, in terms of
timing and configuration of the maximum stages, location and durability of the
major nucleation centres, and patterns of ice retreat during the last
deglaciation.

@@

The exact timing of modelled MIS~2 maximum ice volume depends strongly on the
choice of applied palaeo-temperature record, which allows for a more in-depth
comparison with geological evidence for the timing of maximum Cordilleran ice
sheet extent. In the Puget lowland (Fig.~\ref{fig:locmap}), the LGM advance of
the southern Cordilleran ice sheet margin has been constrained by radiocarbon
dating on wood between 17.4 and 16.4\,\unit{\chem{^{14}C}\,cal\,ka}
\citep{Porter.Swanson.1998}. These dates are consistent with radiocarbon dates
from the offshore sedimentary record, which reveals
an increase of glaciomarine sedimentation between 19.5 and
16.2\,\unit{\chem{^{14}C}\,cal\,ka} \citep{Cosma.etal.2008, Taylor.etal.2014}.
Radiocarbon
dating of the northern Cordilleran ice sheet margin is much less constrained
but straddles presented constraints from the southern margin. However,
cosmogenic exposure dating places the timing of maximum CIS extent during the
McConnell glaciation close to 17\,\unit{\chem{^{10}Be}\,ka}
\citep{Stroeven.etal.2010, Stroeven.etal.2014}. A sharp transition in
the sediment record of the Gulf of Alaska indicates a retreat of regional
outlet glaciers onto land at 14.8\,\unit{\chem{^{14}C}\,cal\,ka}
\citep{Davies.etal.2011}.

Among the simulations presented here, only those forced with the GRIP, EPICA
and Vostok palaeo-temperature records yield Cordilleran ice sheet maximum
extents that may be compatible with these field constraints
(Fig.~\ref{fig:timeseries}, lower panel; Table~\ref{tab:extrema}).
Simulations driven by the NGRIP, ODP~1012 and ODP~1020
palaeo-temperature records, on the contrary, yield MIS~2 maximum Cordilleran
ice sheet extents that pre-date field-based constraints by several thousands of
years (about 6, 6 and 4\,ka respectively).
Concerning the simulations driven by oceanic records, this early
deglaciation is caused by an early warming present in the alkenone
palaeo-temperature reconstructions (Fig.~\ref{fig:timeseries}, upper panel;
\citealp[Fig.~3]{Herbert.etal.2001}). However, this
early warming is a local effect, corresponding to a weakening of the California
current \citep{Herbert.etal.2001}. The California current, driving cold
waters southwards along the south-western coast of North America,
has been shown to have weakened during each peak global glaciation (in SPECMAP)
during the past 550\,ka, including the LGM, resulting in paradoxically warmer
sea-surface temperatures at the locations of the ODP~1012 and ODP~1020 sites
\citep{Herbert.etal.2001}.

Because most of the marine margin of the Cordilleran ice sheet terminated in a
sector of the Pacific Ocean unaffected by variations in the California current,
it probably remained insensitive to this local phenomenum. However, the above
paradox
illustrates the complexity of ice-sheet feedbacks on regional climate, and
demonstrates that, although located in the neighbourhood of the modelling
domain, the ODP~1012 and ODP~1020 palaeo-temperature records cannot be
used as a realistic forcing to model the Cordilleran ice sheet
through the last glacial cycle.

Similarly, the simulation using the NGRIP palaeo-temperature record depicts an
early onset of deglaciation (Fig.~\ref{fig:timeseries}) following its last
glacial volume maximum attained about 6\,ka earlier (22.9\,ka,
Table~\ref{tab:extrema}) than dated evidence of the last glacial maximum
advance. In addition, there exist a fair agreement between the EPICA and Vostok
alaeo-temperature records, resulting in only little differences between the
simulations driven by those records. These differences are not analysed
further; instead we focus the rest of our analysis on simulations forced by
palaeo-temperature records from the GRIP and EPICA ice cores that appear to
produce the most realistic reconstructions of regional glaciation history among
all records tested, yet bearing significant disparities. To allow for a more
detailed comparison against the geologic record, these two simulations were
re-run using the high-resolution grid (Fig.~\ref{fig:timeseries}, lower panel,
dotted lines).

@@

This could reflect the fact that our model does not include
feedback mechanisms between ice sheet topography and the regional climate.
Firstly, during the build-up phase preceding the LGM, rapid accumulation over
the Coast Mountains enhanced the topographic barrier formed by these mountain
ranges, which likely resulted in a decrease of precipitation and, therefore,
a decrease of
accumulation in the interior. Secondly, latent warming of the moisture-depleted
air parcels flowing over this enhanced topography could have resulted in an
inflow of potentially warmer air over the eastern flank of the ice sheet,
increasing melt
along the advancing margin \citep[cf.][]{Langen.etal.2012}. Because these two
processes, both with a tendency to limit ice-sheet growth, are absent from
our model, the
eastern margin of the ice sheet and the position of the main meridional ice
divide are certainly biased towards the east in our simulations
\citep{Seguinot.etal.2014}.

However, field-based palaeoglaciological reconstructions have
struggled to reconcile the more westerly-centred ice divide in south-central
British Columbia with evidence in the Rocky Mountains and beyond, that the
Cordilleran ice sheet invaded the western Interior Plains, where it merged with
the southwestern margin of the Laurentide ice sheet and was deflected to the
south \citep{Jackson.etal.1997, Bednarski.Smith.2007, Kleman.etal.2010,
Margold.etal.2013, Margold.etal.2013a}. Ice geometries from our model runs do
not have this problem, because the position and elevation of the ice divide
results in significant ice drainage across the Rocky Mountains at the LGM
(Fig.~\ref{fig:mis2}).

During MIS~2, total ice volume peaks at 8.01\,m~s.l.e. in the GRIP simulation
and at 8.77\,m~s.l.e. in the EPICA simulation.

@@

The modelled ice sheet configurations corresponding to ice volume maxima during
MIS~4 are more sensitive to the choice of atmospheric forcing than those
corresponding to ice volume maxima during MIS~2 (Fig.~\ref{fig:snapshots},
upper panels;
Fig.~\ref{fig:mis4}). The GRIP simulation (Fig.~\ref{fig:mis4}, left panel)
results in a modelled maximum ice sheet extent that closely resembles that
obtained
during MIS~2, with the only major difference being slightly less extensive
northern and eastern sectors. In contrast, the EPICA simulation produces a lower
ice volume maximum (Fig.~\ref{fig:timeseries}), which translates in the
modelled ice sheet geometry into a significantly reduced southern sector, more
restricted ice cover in northern and eastern sectors, and generally lower
ice surface elevations in the interior (Fig.~\ref{fig:mis4}, left panel).
Thus, only the GRIP simulation can explain the presence of MIS~4 glacial
deposits in the Puget Lowland \citep{Troost.2014} and that of ice-rafted debris
in the marine sediment record offshore Vancouver Island at
ca.~47\,\unit{\chem{^{14}C}\,cal\,ka} \citep{Cosma.etal.2008}.

In the GRIP simulation, some sections of the MIS~4 maximum ice margin are
slightly more extensive than their MIS~2 counterparts. This is notably the case
in the Puget Lowland and along parts of the ice margin in the marine sectors,
on the north slope of the Alaska Range and in Yukon Territory. However, these
differences
are small and generally correspond to a few grid cells only. Although such
results may be corroborated by field evidence in some places
\citep[e.g.][]{Ward.etal.2007}, considering the simplicity of the applied
climate forcing, modelled differences in extent are not demonstrably
significant.

During MIS~4, total ice volume peaks at 7.19\,m~s.l.e. in the GRIP simulation
and at 5.04\,m~s.l.e. in the EPICA simulation, corresponding to respectively
90 and 57\,\% of the MIS~2 ice volumes.

@@

A notable exception to the transient character of other sectors of the
Cordilleran ice sheet
is the northern slope of the Alaska Range, where modelled glaciers are confined
to its foothills during the entire simulation period (Fig.~\ref{fig:duration},
AR). This apparent insensitivity of modelled glacial extent to temperature
fluctuations results from a combination of low precipitation, high summer
temperature and large temperature standard deviation for this sector of the
modelling domain (Fig.~\ref{fig:atm}). This result could potentially explain
the local distribution of glacial deposits, which indicates that these glaciers
have remained small throughout the Pleistocene \citep{Kaufman.Manley.2004}.

@@

It is generally believed that the Cordilleran ice sheet formed by the
coalescence of several mountain-centred ice caps \citep{Davis.Mathews.1944}.
In our simulations, major ice-dispersal centres, visible on the modelled ice
cover duration maps (Fig.~\ref{fig:duration}), are located over the Coast
Mountains (CM), the Columbia and Rocky mountains (CRM), the Skeena Mountains
(SM), and the Selwyn and Mackenzie mountains (SMKM). Although the Coast,
Columbia, Rocky and Skeena mountains (CM, CRM, SM) are covered by mountain
glaciers for most of the last glacial cycle, providing durable nucleation
centres for an ice sheet, this is not the case for the Selwyn and
Mackenzie mountains (SMKM), where ice cover on the highest peaks is limited to a
fraction of the last glacial cycle. In other words, the Selwyn and Mackenzie
mountains appear as a secondary ice-dispersal centre which is only active
during the coldest periods of the last glacial cycle. The Northern Rocky
Mountains (Fig.~\ref{fig:duration}, NRM) do not act as a nucleation centre,
but rather as a pinning point for the Cordilleran ice sheet margin coming from
the west.

Perhaps the most striking feature displayed by the distributions of modelled
ice cover is the persistence of the Skeena Mountains ice cap throughout the
entire last glacial cycle and its predominance over the other ice-dispersal
centres
(Figs.~\ref{fig:snapshots} and~\ref{fig:duration}, SM). Regardless of the applied
forcing, this ice cap appears to survive MIS~3 (Fig.~\ref{fig:snapshots},
middle panels), and serves as a nucleation centre at the onset of the glacial
readvance towards the LGM (MIS~2). This situation appear similar to the neighbouring Laurentide ice sheet, for which the importance of residual ice for
glacial history leading up to the LGM has been illustrated by the
MIS~3 residual ice bodies in northern and eastern Canada as nucleation centres
for a much more extensive MIS~2 configuration \citep{Kleman.etal.2010}.

The presence of a Skeena Mountains ice cap during most of the last glacial
cycle can be explained by a currently more eastwards-spread distribution of
winter precipitation there than along other transects of the
modelling domain (Fig.~\ref{fig:atm}). Along most of the north-western coast of
North America, coastal mountain ranges form a pronounced topographic barrier
for westerly winds, capturing atmospheric moisture in the form of orographic
precipitation, and resulting in arid interior lowlands. However, near the
centre of our modelling domain, this barrier is less pronounced than elsewhere,
allowing westerly winds to carry moisture
further inland, until it is captured by the extensive Skeena Mountains in
north-central British Columbia, thus resulting in a more widespread
distribution of winter precipitation (Fig.~\ref{fig:atm}).

During MIS~3, total ice volume corresponding to these persistent ice-dispersal
centres reaches a minimum of 1.52\,m~s.l.e. in the GRIP simulation and of
2.68\,m~s.l.e. in the EPICA simulation, corresponding to respectively 19 and
31\,\% of the MIS~2 ice volumes.

@@

The modelled distribution of warm-based ice cover (Figs.~\ref{fig:warmbase}
and~\ref{fig:warmfrac}) is inevitably affected by our assumption of a constant,
70\,\unit{mW\,m^{-2}} geothermal heat flux at 3\,km depth
(Sect.~\ref{sec:icedyn}). However, both the Skeena Mountains and the area
west of the Mackenzie Mountains are areas of higher-than-average geothermal
heat flux \citep{Blackwell.Richards.2004}. We can therefore expect longer
durations of warm-based ice cover for these areas if we were to include
spatially variable geothermal forcing in our Cordilleran ice sheet simulations.

@@

Similarly to other glaciated regions, most glacial traces in the North American
Cordillera relate to the last few millennia of glaciation, because most of the
older evidence has been overprinted by warm-based ice retreat
during the last deglaciation \citep{Kleman.1994, Kleman.etal.2010}. From a
numerical modelling perspective, phases of glacier retreat are more challenging
than phases of growth, because they involve more rapid fluctuations of the ice
margin, increased flow velocities and longitudinal stress gradients, and poorly
understood hydrological processes. The latter are typically included in the
models through simple parametrisations \citep[e.g.][]{Clason.etal.2012,
Clason.etal.2014, Bueler.Pelt.2014}, if included at all. However, next after a
matching of modelled ice margin positions with mapped end moraines
\citep{Napieralski.etal.2007}, indicative of maximum ice sheet extents
(Sect.~\ref{sec:mis2} and~\ref{sec:mis4}), geomorphologically-based
reconstructions of patterns of of ice sheet retreat during the last
deglaciation provide the second best source of evidence for the validation of
our simulations.

@@

In our simulations, the timing of peak ice volume during the last glacial
maximum and the pacing of deglaciation depend critically on the choice of
climate forcing (Table~\ref{tab:extrema}; Figs.~\ref{fig:timeseries}
and~\ref{fig:deglacseries}). Adopting the EPICA climate forcing yields peak ice
volume at 17.2\,ka and an uninterrupted deglaciation until about 9\,ka
(Fig.~\ref{fig:deglacseries}, lower panel, red curves). On the contrary, the
simulation driven by the GRIP palaeo-temperature record yields peak ice volume
at 19.3\,ka and a deglaciation interrupted by two ice volume standstills until
about 8\,ka. The first interruption occurs between 16.6 and 14.5\,ka, and the
second between 13.1 and 11.6\,ka (Fig.~\ref{fig:deglacseries}, lower panel,
blue curve).

Hence, the two model runs, while similar in overall timing compared to runs
with other climate drivers, differ in detail. On the one hand, the EPICA run
depicts peak glaciation about 2\,ka later than the GRIP run, in closer
agreement with dated maximum extents, and shows a faster, uninterrupted
deglaciation which yields sporadic ice cover more than 1\,ka earlier. On the
other hand, the GRIP run yields a deglaciation in three steps, compatible with
marine sediment sequences offshore Vancouver Island, where the distribution of
ice-rafted debris indicate an ice margin retreat from the Georgia Strait in two
phases that are contemporary with warming oceanic temperatures from 17.2 to
16.5 and from 15.5 to 14.0\,\unit{\chem{^{14}C}\,cal\,ka}
\citep{Taylor.etal.2014}.

>>>>

% Running title
\runningtitle{Numerical simulations of the Cordilleran ice sheet
              through the last glacial cycle}

@@

% =============================================================================
\section{Comparison with the geologic record}
\label{sec:discussion}
% =============================================================================

Large variations in the model responses to evolving climate forcing reveal its
sensitivity to the choice of palaeo-temperature proxy record. To distinguish
between different records, geological evidence of former glaciations
provide a basis for validation of our runs, while the results from numerical
modelling can perhaps help to analyse some of the complexity of this evidence.
In this section, we compare model outputs to the geologic record, in terms of
timing and configuration of the maximum stages, location and lifetime of
major nucleation centres, and patterns of ice retreat during the last
deglaciation.

@@

The exact timing of modelled MIS~2 maximum ice volume depends strongly on the
choice of applied palaeo-temperature record, which allows for a more in-depth
comparison with geological evidence for the timing of maximum Cordilleran ice
sheet extent. In the Puget Lowland (Fig.~\ref{fig:locmap}), the LGM advance of
the southern Cordilleran ice sheet margin has been constrained by radiocarbon
dating on wood between 17.4 and 16.4\,\unit{\chem{^{14}C}\,cal\,ka}
\citep{Porter.Swanson.1998}. These dates are consistent with radiocarbon dates
from the offshore sedimentary record, which reveals
an increase of glaciomarine sedimentation between 19.5 and
16.2\,\unit{\chem{^{14}C}\,cal\,ka} \citep{Cosma.etal.2008, Taylor.etal.2014}.
Radiocarbon
dating of the northern Cordilleran ice sheet margin is much less constrained
but straddles presented constraints from the southern margin. However,
cosmogenic exposure dating places the timing of maximum CIS extent during the
McConnell glaciation close to 17\,\unit{\chem{^{10}Be}\,ka}
\citep{Stroeven.etal.2010, Stroeven.etal.2014}. A sharp transition in
the sediment record of the Gulf of Alaska indicates a retreat of regional
outlet glaciers onto land at 14.8\,\unit{\chem{^{14}C}\,cal\,ka}
\citep{Davies.etal.2011}.

Among the simulations presented here, only those forced with the GRIP, EPICA
and Vostok palaeo-temperature records yield Cordilleran ice sheet maximum
extents that may be compatible with these field constraints
(Fig.~\ref{fig:timeseries}, lower panel; Table~\ref{tab:extrema}).
Simulations driven by the NGRIP, ODP~1012 and ODP~1020
palaeo-temperature records, on the contrary, yield MIS~2 maximum Cordilleran
ice sheet volumes that pre-date field-based constraints by several thousands of
years (about 6, 6 and 4\,ka respectively).
Concerning the simulations driven by oceanic records, this early
deglaciation is caused by an early warming present in the alkenone
palaeo-temperature reconstructions (Fig.~\ref{fig:timeseries}, upper panel;
\citealp[Fig.~3]{Herbert.etal.2001}). However, this
early warming is a local effect, corresponding to a weakening of the California
current \citep{Herbert.etal.2001}. The California current, driving cold
waters southwards along the south-western coast of North America, has been
shown to have weakened during each peak of global glaciation (in SPECMAP)
during the past 550\,ka, including the LGM, resulting in paradoxically warmer
sea-surface temperatures at the locations of the ODP~1012 and ODP~1020 sites
\citep{Herbert.etal.2001}.

Because most of the marine margin of the Cordilleran ice sheet terminated in a
sector of the Pacific Ocean unaffected by variations in the California current,
it probably remained insensitive to this local phenomenum. However, the above
paradox
illustrates the complexity of ice-sheet feedbacks on regional climate, and
demonstrates that, although located in the neighbourhood of the modelling
domain, the ODP~1012 and ODP~1020 palaeo-temperature records cannot be
used as a realistic forcing to model the Cordilleran ice sheet
through the last glacial cycle.
Similarly, the simulation using the NGRIP palaeo-temperature record depicts an
early onset of deglaciation (Fig.~\ref{fig:timeseries}) following its last
glacial volume maximum (22.9\,ka, Table~\ref{tab:extrema}) attained about 6\,ka
earlier than dated evidence of the LGM advance.
There exist a fair agreement between the EPICA and Vostok
alaeo-temperature records, resulting in only little differences between the
simulations driven by those records. These differences are not analysed
further; instead we focus the rest of our analysis on simulations forced by
palaeo-temperature records from the GRIP and EPICA ice cores that appear to
produce the most realistic reconstructions of regional glaciation history among
all records tested, yet bearing significant disparities in model output. To
allow for a more detailed comparison against the geological record, these two
simulations were re-run using a higher-resolution grid (Sect.~\ref{sec:model};
Fig.~\ref{fig:timeseries}, lower panel, dotted lines).

@@

Such deviation from the geological inferences could reflect the fact that our
model does not include
feedback mechanisms between ice sheet topography and the regional climate.
Firstly, during the build-up phase preceding the LGM, rapid accumulation over
the Coast Mountains enhanced the topographic barrier formed by these mountain
ranges, which likely resulted in a decrease of precipitation and, therefore,
a decrease of
accumulation in the interior. Secondly, latent warming of the moisture-depleted
air parcels flowing over this enhanced topography could have resulted in an
inflow of potentially warmer air over the eastern flank of the ice sheet,
increasing melt
along the advancing margin \citep[cf.][]{Langen.etal.2012}. Because these two
processes, both with a tendency to limit ice-sheet growth, are absent from
our model, the
eastern margin of the ice sheet and the position of the main meridional ice
divide are certainly biased towards the east in our simulations
\citep{Seguinot.etal.2014}.

However, field-based palaeoglaciological reconstructions have
struggled to reconcile the more westerly-centred ice divide in south-central
British Columbia with evidence in the Rocky Mountains and beyond, that the
Cordilleran ice sheet invaded the western Interior Plains, where it merged with
the southwestern margin of the Laurentide ice sheet and was deflected to the
south \citep{Jackson.etal.1997, Bednarski.Smith.2007, Kleman.etal.2010,
Margold.etal.2013, Margold.etal.2013a}. Ice geometries from our model runs do
not have this problem, because the position and elevation of the ice divide
ensure significant ice drainage across the Rocky Mountains at the LGM
(Fig.~\ref{fig:mis2}).

During MIS~2, the modelled total ice volume peaks at 8.01\,m~s.l.e. in the GRIP
simulation
and at 8.77\,m~s.l.e. in the EPICA simulation.

@@

The modelled ice sheet configurations corresponding to ice volume maxima during
MIS~4 are more sensitive to the choice of atmospheric forcing than those
corresponding to ice volume maxima during MIS~2 (Fig.~\ref{fig:snapshots},
upper panels;
Fig.~\ref{fig:mis4}). The GRIP simulation (Fig.~\ref{fig:mis4}, left panel)
results in a modelled maximum ice sheet extent that closely resembles that
obtained during MIS~2, with the only major difference of being slightly less
extensive across
northern and eastern sectors. In contrast, the EPICA simulation produces a lower
ice volume maximum (Fig.~\ref{fig:timeseries}), which translates in the
modelled ice sheet geometry into a significantly reduced southern sector, more
restricted ice cover in northern and eastern sectors, and generally lower
ice surface elevations in the interior (Fig.~\ref{fig:mis4}, left panel).
Thus, only the GRIP simulation can explain the presence of MIS~4 glacial
deposits in the Puget Lowland \citep{Troost.2014} and that of ice-rafted debris
in the marine sediment record offshore Vancouver Island at
ca.~47\,\unit{\chem{^{14}C}\,cal\,ka} \citep{Cosma.etal.2008}.

During MIS~4, the modelled total ice volume peaks at 7.19\,m~s.l.e. in the GRIP
simulation
and at 5.04\,m~s.l.e. in the EPICA simulation, corresponding to respectively
90 and 57\,\% of the MIS~2 ice volumes.

@@

A notable exception to the transient character of the maximum extent of
Cordilleran ice sheet
is the northern slope of the Alaska Range, where modelled glaciers are confined
to its foothills during the entire simulation period (Fig.~\ref{fig:duration},
AR). This apparent insensitivity of modelled glacial extent to temperature
fluctuations results from a combination of low precipitation, high summer
temperature and large temperature standard deviation in the plains of the
Alaska Interior (Fig.~\ref{fig:atm}) which confines glaciation to the
foothills of the mountains. This result could potentially explain
the local distribution of glacial deposits, which indicates that glaciers
flowing on the northern slope of the Alaska Range
have remained small throughout the Pleistocene \citep{Kaufman.Manley.2004}.

@@

It is generally believed that the Cordilleran ice sheet formed by the
coalescence of several mountain-centred ice caps \citep{Davis.Mathews.1944}.
In our simulations, major ice-dispersal centres, visible on the modelled ice
cover duration maps (Fig.~\ref{fig:duration}), are located over the Coast
Mountains (CM), the Columbia and Rocky mountains (CRM), the Skeena Mountains
(SM), and the Selwyn and Mackenzie mountains (SMKM). The Wrangell and
St.-Elias mountains, heavily glacierized at present, host an ice cap for the
entire length of both simulations, but that ice cap does not appear to be a
major feed to the Cordilleran ice sheet (Fig.~\ref{fig:duration}, WSEM).
Although the Coast,
Columbia, Rocky and Skeena mountains (CM, CRM, SM) are covered by mountain
glaciers for most of the last glacial cycle, providing durable nucleation
centres for an ice sheet, this is not the case for the Selwyn and
Mackenzie mountains (SMKM), where ice cover on the highest peaks is limited to a
small fraction of the last glacial cycle. In other words, the Selwyn and Mackenzie
mountains appear as a secondary ice-dispersal centre which is only active
during the coldest periods of the last glacial cycle. The Northern Rocky
Mountains (Fig.~\ref{fig:duration}, NRM) do not act as a nucleation centre,
but rather as a pinning point for the Cordilleran ice sheet margin coming from
the west.

Perhaps the most striking feature displayed by the distributions of modelled
ice cover is the persistence of the Skeena Mountains ice cap throughout the
entire last glacial period (ca.~100--10\,ka) and its predominance over the
other ice-dispersal centres
(Figs.~\ref{fig:snapshots} and~\ref{fig:duration}, SM). Regardless of the applied
forcing, this ice cap appears to survive MIS~3 (Fig.~\ref{fig:snapshots},
middle panels), and serves as a nucleation centre at the onset of the glacial
readvance towards the LGM (MIS~2). This situation appear similar to the neighbouring Laurentide ice sheet, for which the importance of residual ice for the
glacial history leading up to the LGM has been illustrated by the
MIS~3 residual ice bodies in northern and eastern Canada as nucleation centres
for a much more extensive MIS~2 configuration \citep{Kleman.etal.2010}.

The presence of a Skeena Mountains ice cap during most of the last glacial
cycle can be explained by a more eastwards-spread distribution of
winter precipitation there than along other transects of the
modelling domain (Fig.~\ref{fig:atm}). Along most of the north-western coast of
North America, coastal mountain ranges form a pronounced topographic barrier
for westerly winds, capturing atmospheric moisture in the form of orographic
precipitation, and resulting in arid interior lowlands. However, near the
centre of our modelling domain, this barrier is less pronounced than elsewhere,
allowing westerly winds to carry moisture
further inland, until it is captured by the extensive Skeena Mountains in
north-central British Columbia, thus resulting in a more widespread
distribution of winter precipitation (Fig.~\ref{fig:atm}).

During MIS~3, the modelled total ice volume corresponding to these persistent
ice-dispersal
centres reaches a minimum of 1.52\,m~s.l.e. in the GRIP simulation and of
2.68\,m~s.l.e. in the EPICA simulation, corresponding to respectively 19 and
31\,\% of the MIS~2 ice volumes.

@@

The modelled distribution of warm-based ice cover (Figs.~\ref{fig:warmbase}
and~\ref{fig:warmfrac}) is inevitably affected by our assumption of a constant,
70\,\unit{mW\,m^{-2}} geothermal heat flux at 3\,km depth
(Sect.~\ref{sec:icedyn}). However, the Skeena Mountains and the area west of
the Mackenzie Mountains experience higher-than-average geothermal heat flux
with measured values of ca.~80 and ca.~100\,\unit{mW\,m^{-2}}
\citep{Blackwell.Richards.2004}. We can therefore expect even longer
durations of warm-based ice cover for these areas if we were to include
spatially variable geothermal forcing in our Cordilleran ice sheet simulations.

@@

Similarly to other glaciated regions, most glacial traces in the North American
Cordillera relate to the last few millennia of glaciation, because most of the
older evidence has been overprinted by warm-based ice retreat
during the last deglaciation \citep{Kleman.1994, Kleman.etal.2010}. From a
numerical modelling perspective, phases of glacier retreat are more challenging
than phases of growth, because they involve more rapid fluctuations of the ice
margin, increased flow velocities and longitudinal stress gradients, and poorly
understood hydrological processes. The latter are typically included in the
models through simple parametrisations \citep[e.g.][]{Clason.etal.2012,
Clason.etal.2014, Bueler.Pelt.2014}, if included at all. However, next after
the mapping of maximum ice sheet extents during MIS~2 and MIS~4
(Sect.~\ref{sec:mis2} and~\ref{sec:mis4}), geomorphologically-based
reconstructions of patterns of ice sheet retreat during the last
deglaciation provide the second best source of evidence for the validation of
our simulations.

@@

In our simulations, the timing of peak ice volume during the LGM
and the pacing of deglaciation depend critically on the choice of
climate forcing (Table~\ref{tab:extrema}; Figs.~\ref{fig:timeseries}
and~\ref{fig:deglacseries}). Adopting the EPICA climate forcing yields peak ice
volume at 17.2\,ka and an uninterrupted deglaciation until about 9\,ka
(Fig.~\ref{fig:deglacseries}, lower panel, red curves). On the contrary, the
simulation driven by the GRIP palaeo-temperature record yields peak ice volume
at 19.3\,ka and a deglaciation interrupted by two phases of regrowth until
about 8\,ka. The first interruption occurs between 16.6 and 14.5\,ka, and the
second between 13.1 and 11.6\,ka (Fig.~\ref{fig:deglacseries}, lower panel,
blue curve).

Hence, the two model runs, while similar in overall timing compared to the runs
with other climate drivers, differ in detail. On the one hand, the EPICA run
depicts peak glaciation about 2\,ka later than the GRIP run, in closer
agreement with dated maximum extents, and shows a faster, uninterrupted
deglaciation which yields sporadic ice cover more than 1\,ka earlier. On the
other hand, the GRIP run yields a deglaciation in three steps, compatible with
marine sediment sequences offshore Vancouver Island, where the distribution of
ice-rafted debris indicates an ice margin retreat from the Georgia Strait in two
phases that are contemporary with warming oceanic temperatures from 17.2 to
16.5 and from 15.5 to 14.0\,\unit{\chem{^{14}C}\,cal\,ka}
\citep{Taylor.etal.2014}.
